---
title: "Modelling Match State and Player Performance in the BBL"
author: "Hieu (Mark), Pranav, Prem"
format: revealjs
editor: visual
---

```{r}
#| echo: FALSE

knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)

library(tidyverse)
library(here)
library(yardstick)
library(xgboost)
```

## Big Bash League (BBL) T20 cricket project {.smaller}

- **Project focus:** Context-aware player evaluation using 14 seasons of BBL ball-by-ball data (2011–2025) to quantify performance under match pressure.

- **Target stakeholders:** BBL franchise coaches & strategy analysts, sports media/broadcasters, and high-performance talent recruiters.

- **Key deliverables:** Interactive exploration dashboard (Shiny), predictive delivery-outcome models, and a contextual performance report.

- **Progress to date:** Full delivery-level match-state feature pipeline built (RRR, phase tags, wickets in hand); initial dashboard and model prototypes.

## Key question {.smaller}

1)  Does winning the toss and choosing to field (the more common decision in T20) actually translate into a meaningful advantage?

2)  Do players naturally cluster into distinct batting or bowling archetypes based on their ball-by-ball performance profiles?

3)  How do scoring rates and wicket probabilities vary across the three phases of a T20 innings — the Powerplay (overs 1–6), the middle overs (7–16), and the death overs (17–20)?

4)  Can we build a model that predicts the probability of a wicket falling on any given delivery, using match-state and player-identity features?

5)  Which batters perform best under high-pressure chase situations — defined by a high required run rate and few wickets in hand — relative to their performance in low-pressure situations?

6)  Which bowlers are most effective at restricting runs and taking wickets in the death overs, after controlling for the match state they bowl in?

## Does winning the toss and choosing to field (the more common decision in T20) actually translate into a meaningful advantage?

<br>

Approach: Use a logistic regression model to test whether winning the toss — and specifically choosing to field — has a real effect on match outcomes.

------------------------------------------------------------------------

``` r
coin_toss_advantage <- bbl_match_info |>
  select(clean_stadium, home_team, toss_winner, toss_decision, winner) |>
  mutate(
    toss_winner_is_home_team = home_team == toss_winner,
    .after = toss_winner
  ) |>
  mutate(winner_is_home_team = home_team == winner) |>
  mutate(winner_same_as_toss_winner = winner == toss_winner) |>
  mutate(across(where(is.character), as.factor)) |>
  mutate(
    # Lumps any stadium with fewer than 10 matches into "Other"
    clean_stadium_grouped = fct_lump_min(as.factor(clean_stadium), 
                                         min = 10, 
                                         other_level = "Other")
  )
```

```r
log_reg <- glm(
  winner_same_as_toss_winner ~
    clean_stadium_grouped + toss_decision + toss_winner,
  coin_toss_advantage,
  family = binomial
)
```

------------------------------------------------------------------------

##  {.scrollable}

::: {style="font-size: 60%;"}
```{r}
log_reg <- readRDS("Mark/log_reg_model.rds")
sjPlot::tab_model(log_reg)
```
:::

```{r}
results <- readRDS(
  here("Pranav", "data", "model_2_results.rds")
)

t20_final <- results$model
t20_cv <- results$cv
test_data <- results$test_data
pred_probs <- results$pred_probs
y_test <- results$y_test
pr_auc <- results$pr_auc
roc_auc <- results$roc_auc
```

## Can we build a model that predicts the probability of a wicket falling on any given delivery, using match-state and player-identity features?

---

### Feature engineering
The variables engineereed for the model are:

#### Match state:

  "innings", "over_no", "phase", "balls_remaining",
  "score_before", "wickets_lost_before", "required_runs", 
  "free_hit",

#### Bowler in-match form:

  "bowler_balls_bowled_before", "bowler_economy_before",
  "bowler_boundary_p_before", "bowler_dot_ball_p_before",

---

#### Batsman in-match form:

  "batsman_runs_scored_before", "batsman_balls_faced_before",
  "batsman_strike_rate_before", "batsman_boundary_p_before",

#### Bistorical / career form:

  "career_matches_before", "career_batting_avg_before",
  "career_bowling_strike_rate_before","career_batting_strike_rate_before", "career_boundary_p_before",
  "career_matches_bowled_before", "career_economy_before",
  "career_bowling_avg_before", "career_dot_p_before"

---

### Training and test split
```r
# Identify the chronological order of matches
match_dates <- model_data |>
  distinct(match_id, start_date) |>
  arrange(start_date, match_id)

# Use the first 80% of matches chronologically as the training period
cutoff_index <- floor(nrow(match_dates) * 0.8)

# Set the date of the 80th percentile match as the cutoff
cutoff_date <- match_dates$start_date[cutoff_index]

# Train the model only on matches occurring before or on the cutoff date
train_data <- model_data |>
  filter(start_date <= cutoff_date)

# Test the model on the most recent 20% of matches
test_data <- model_data |>
  filter(start_date > cutoff_date)
```

---

### Model

An XGBoost model was used on the data. This initial model had a PR-AUC score of 0.096, and an ROC-AUC score of 0.623. 

---

### Feature importance

```{r}
importance <- xgb.importance(
  model = t20_final
)

importance |>
  slice_max(
    order_by = Gain,
    n = 10
  ) |>
  ggplot(
    aes(
      x = reorder(Feature, Gain),
      y = Gain
    )
  ) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Predictors",
    x = NULL,
    y = "XGBoost Gain"
  ) +
  theme_minimal()
```

This shows that the initial iterations of the model are giving more importance to match-level variables and not career-level variables. 

---

### Next Steps

1. Tune and validate the wicket model : Test different XGBoost parameters and compare performance against simpler baseline models.

2. Analyse model errors : Identify match situations where the model struggles to predict wickets.

3. Answer the next question : "Which bowlers are most effective at restricting runs and taking wickets in the death overs, after controlling for the match state they bowl in?"

## How do scoring rates and wicket probabilities vary across the three phases of a T20 innings — the Powerplay (overs 1–6), the middle overs (7–16), and the death overs (17–20)?

---

Dashboard idea using cricketR -

1.  Perform player specific statistic using existing functions -

- batsmanCumulativeAverageRuns() - batsmanCumulativeStrikeRate() - work on it and add it

batsmanMovingAverage(), batsmanPerfBoxHist() - good graph for keeping a track of player performance over the season and comparing who scores at which point

------------------------------------------------------------------------

```r
library(shiny)
library(dplyr)
library(plotly)

# -----------------------------
# UI
# -----------------------------

ui <- fluidPage(
  
  titlePanel("BBL Match Scoring Analysis"),
  
  sidebarLayout(
    
    sidebarPanel(
      
      selectInput(
        inputId = "season",
        label = "Select Season:",
        choices = unique(over_data$season)
      ),
      
      selectInput(
        inputId = "match",
        label = "Select Match:",
        choices = NULL
      )
      
    ),
    
    mainPanel(
      
      plotlyOutput(
        outputId = "score_plot",
        height = "600px"
      )
      
    )
  )
)


# -----------------------------
# SERVER
# -----------------------------

server <- function(input, output, session) {
  
  
  # Update available matches when season changes
  observeEvent(input$season, {
    
    matches <- over_data |>
      filter(season == input$season) |>
      distinct(match_id) |>
      pull(match_id)
    
    updateSelectInput(
      session,
      inputId = "match",
      choices = matches
    )
    
  })
  
  
  # Filter selected match
  selected_match <- reactive({
    
    req(input$season, input$match)
    
    over_data |>
      filter(
        season == input$season,
        match_id == input$match
      )
    
  })
  
  
  # Interactive scoring comparison graph
  output$score_plot <- renderPlotly({
    
    match_data <- selected_match()
    
    plot_ly(
      data = match_data,
      x = ~over,
      y = ~cumulative_runs,
      color = ~batting_team,
      type = "scatter",
      mode = "lines+markers",
      
      text = ~paste0(
        "Team: ", batting_team,
        "<br>Over: ", over,
        "<br>Score: ", score,
        "<br>Runs this over: ", runs_in_over,
        "<br>Wickets this over: ", wickets_in_over,
        "<br>Phase: ", Phase_over
      ),
      
      hoverinfo = "text"
    ) |>
      
      layout(
        
        title = "Scoring Comparison",
        
        xaxis = list(
          title = "Over",
          dtick = 1,
          range = c(1, 20)
        ),
        
        yaxis = list(
          title = "Cumulative Runs"
        ),
        
        legend = list(
          title = list(
            text = "Team"
          )
        )
      )
    
  })
  
}


# -----------------------------
# RUN APP
# -----------------------------

shinyApp(ui = ui, server = server)
```