---
title: "Untitled"
format: html
---

```{r Loading Libraries}
library(tidyverse)
```

```{r Reading in Data}
bbl_match_info <-  read_csv("./data/bbl_match_info.csv")
bbl_all <- read_csv("./data/bbl_all.csv")
```


```{r}
summary(bbl_match_info)
glimpse(bbl_all)
```

## 1. Creating Necessary Variables

To check how legal balls are recorded, let us check up on the rows pertaining to deliveries


```{r}
bbl_delivery <- bbl_all|> select(ball,actual_delivery,wides, noballs,extras)
view(bbl_delivery)
bbl_delivery|> count(extras)
```

Creating a runnning score and per delivery score
```{r}
bbl_all <- bbl_all|> group_by(match_id,innings)|>
    mutate(delivery_runs = runs_off_bat+extras,
    current_score = cumsum(delivery_runs),
    score_before=lag((current_score),default=0))
```

Creating dot ball indicator
```{r}
bbl_all <- bbl_all |>mutate(
    dot_ball = ifelse(runs_off_bat == 0 & extras == 0, 1, 0)
  )

```

Creating over number and phase of play

```{r}
bbl_all <- bbl_all|> mutate(over_no = floor(ball)+1,
phase= case_when(over_no >=1 & over_no <=6 ~ "Powerplay",
over_no >=7 & over_no <=15 ~ "Middle",
over_no >=16 & over_no <=20 ~ "Death")) 
```


Now to determine legal balls and remaining balls

```{r}
bbl_all <-  bbl_all|> mutate(legal_ball = ifelse(is.na(noballs) & is.na(wides),1,0),
balls_remaining= lag(120-cumsum(legal_ball),default=120))

```

Calculating target and runs required

```{r}
bbl_all <-  bbl_all|> group_by(match_id)|>
    mutate(target=max(current_score[innings==1])+1,
    required_runs=ifelse(innings==2,(target-score_before),NA_real_))
```

Calculating wickets lost by batting team

```{r}
bbl_all <-  bbl_all|> group_by(match_id, innings)|>
    mutate(wickets = ifelse(is.na(wicket_type),0,1) ,
    wickets_lost_before = cumsum(lag(wickets, default = 0)))
```

Creating a boundary indicator

```{r}
bbl_all <-  bbl_all|> group_by(match_id, innings)|>
    mutate(boundary_type=case_when(runs_off_bat== 4 & is.na(non_boundary)  ~ "Four",
                                    runs_off_bat== 6 & is.na(non_boundary)  ~ "Six",
                                    TRUE ~ NA), 
boundary_runs = ifelse(runs_off_bat %in% c(4,6),runs_off_bat,0))

```

Free hit indicator(useful for modelling, no concern to other metrics)

```{r}
bbl_all <- bbl_all |>group_by(match_id, innings) |>
  mutate(
    free_hit = ifelse(
      lag(noballs) > 0,
      1,
      0 ))
```

Now we will create bowler economy:

```{r}
bbl_all <-  bbl_all|> group_by(match_id,bowler)|>
    mutate(bowler_balls_bowled_before = 
                      lag(cumsum(legal_ball),default = 0),

            runs_conceded_before =
                     lag(cumsum(runs_off_bat+ coalesce(wides,0) + coalesce(noballs,0)),
                     default = 0),

            bowler_economy_before = ifelse(bowler_balls_bowled_before > 0 ,
                     (runs_conceded_before/bowler_balls_bowled_before)*6,
                     NA_real_))
                     
```

Another statistic is boundary % per bowler.

```{r}
bbl_all <- bbl_all|> group_by(match_id,bowler)|>
    mutate(bowler_boundary_p_before = ifelse(runs_conceded_before>0,
                lag(cumsum(boundary_runs),default=0)/runs_conceded_before*100,
                0
    ))

```


```{r}
bbl_all <- bbl_all |>
  group_by(match_id, bowler) |>
  mutate(
    bowler_dot_ball_p_before = ifelse(
      bowler_balls_bowled_before > 0,
      lag(cumsum(dot_ball), default = 0) /
        bowler_balls_bowled_before * 100,
      NA_real_ ))
      
```

BATSMAN VARIABLES:

Cumulative runs(lagged)
```{r}
bbl_all <- bbl_all|> group_by(match_id,striker)|>
    mutate(batsman_runs_scored_before= cumsum(lag(runs_off_bat,default=0)))
```


```{r}
bbl_all <- bbl_all |>
  group_by(match_id, striker) |>
  mutate(
 
    batsman_balls_faced_before = cumsum(
      lag(ifelse(is.na(wides), 1, 0), default = 0) #as wides will not be counted
    ),
    
    batsman_strike_rate_before = ifelse(
      batsman_balls_faced_before > 0,
      batsman_runs_scored_before /
        batsman_balls_faced_before * 100,
      NA_real_
    ),
    
    batsman_boundary_p_before = ifelse(
      batsman_runs_scored_before > 0,
      lag(cumsum(boundary_runs), default = 0) /
        batsman_runs_scored_before * 100,
      NA_real_
    )
  )
```