---
title: "Untitled"
format: html
---

```{r Loading Libraries}
library(tidyverse)
```

```{r Reading in Data}
bbl_match_info <-  read_csv("./data/bbl_match_info.csv")
bbl_all <- read_csv("./data/bbl_all.csv")
```


```{r}
summary(bbl_match_info)
glimpse(bbl_all)
```

## 1. Creating Necessary Variables

To check how legal balls are recorded, let us check up on the rows pertaining to deliveries


```{r}
bbl_delivery <- bbl_all|> select(ball,actual_delivery,wides, noballs,extras)
view(bbl_delivery)
bbl_delivery|> count(extras)
```

Creating a runnning score and per delivery score
```{r}
bbl_all <- bbl_all|> group_by(match_id,innings)|>
    mutate(delivery_runs = runs_off_bat+extras,
    current_score = cumsum(delivery_runs),
    score_before=lag((current_score),default=0))
```

Creating dot ball indicator
```{r}
bbl_all <- bbl_all |>mutate(
    dot_ball = ifelse(runs_off_bat == 0 & extras == 0, 1, 0)
  )

```

Creating over number and phase of play

```{r}
bbl_all <- bbl_all|> mutate(over_no = floor(ball)+1,
phase= case_when(over_no >=1 & over_no <=6 ~ "Powerplay",
over_no >=7 & over_no <=15 ~ "Middle",
over_no >=16 & over_no <=20 ~ "Death")) 
```


Now to determine legal balls and remaining balls

```{r}
bbl_all <-  bbl_all|> mutate(legal_ball = ifelse(is.na(noballs) & is.na(wides),1,0),
balls_remaining= lag(120-cumsum(legal_ball),default=120))

```

Calculating target and runs required

```{r}
bbl_all <-  bbl_all|> group_by(match_id)|>
    mutate(target=max(current_score[innings==1])+1,
    required_runs=ifelse(innings==2,(target-score_before),NA_real_))
```

Calculating wickets lost by batting team

```{r}
bbl_all <-  bbl_all|> group_by(match_id, innings)|>
    mutate(wickets = ifelse(is.na(wicket_type),0,1) ,
    wickets_lost_before = cumsum(lag(wickets, default = 0)))
```

Creating a boundary indicator

```{r}
bbl_all <-  bbl_all|> group_by(match_id, innings)|>
    mutate(boundary_type=case_when(runs_off_bat== 4 & is.na(non_boundary)  ~ "Four",
                                    runs_off_bat== 6 & is.na(non_boundary)  ~ "Six",
                                    TRUE ~ NA), 
boundary_runs = ifelse(runs_off_bat %in% c(4,6),runs_off_bat,0))

```

Free hit indicator(useful for modelling, no concern to other metrics)

```{r}
bbl_all <- bbl_all |>group_by(match_id, innings) |>
  mutate(
    free_hit = ifelse(
      lag(noballs) > 0,
      1,
      0 ))
```

Now we will create bowler economy:

```{r}
bbl_all <-  bbl_all|> group_by(match_id,bowler)|>
    mutate(bowler_balls_bowled_before = 
                      lag(cumsum(legal_ball),default = 0),

            runs_conceded_before =
                     lag(cumsum(runs_off_bat+ coalesce(wides,0) + coalesce(noballs,0)),
                     default = 0),

            bowler_economy_before = ifelse(bowler_balls_bowled_before > 0 ,
                     (runs_conceded_before/bowler_balls_bowled_before)*6,
                     NA_real_))
                     
```

Another statistic is boundary % per bowler.

```{r}
bbl_all <- bbl_all|> group_by(match_id,bowler)|>
    mutate(bowler_boundary_p_before = ifelse(runs_conceded_before>0,
                lag(cumsum(boundary_runs),default=0)/runs_conceded_before*100,
                0
    ))

```


```{r}
bbl_all <- bbl_all |>
  group_by(match_id, bowler) |>
  mutate(
    bowler_dot_ball_p_before = ifelse(
      bowler_balls_bowled_before > 0,
      lag(cumsum(dot_ball), default = 0) /
        bowler_balls_bowled_before * 100,
      NA_real_ ))
      
```

BATSMAN VARIABLES:

Cumulative runs(lagged)
```{r}
bbl_all <- bbl_all|> group_by(match_id,striker)|>
    mutate(batsman_runs_scored_before= cumsum(lag(runs_off_bat,default=0)))
```


```{r}
bbl_all <- bbl_all |>
  group_by(match_id, striker) |>
  mutate(
 
    batsman_balls_faced_before = cumsum(
      lag(ifelse(is.na(wides), 1, 0), default = 0) #as wides will not be counted
    ),
    
    batsman_strike_rate_before = ifelse(
      batsman_balls_faced_before > 0,
      batsman_runs_scored_before /
        batsman_balls_faced_before * 100,
      NA_real_
    ),
    
    batsman_boundary_p_before = ifelse(
      batsman_runs_scored_before > 0,
      lag(cumsum(boundary_runs), default = 0) /
        batsman_runs_scored_before * 100,
      NA_real_
    )
  )
```


HISTORICAL PLAYER STATISTICS
For historical stats we have to make sure that the bowler was credited for the particular dismissal(this is not always the case as retired out is not counted) and that the ball faced by the batsman is not a wide.

```{r}
bbl_all <- bbl_all |>
  group_by(match_id, bowler) |>
  mutate(
    bowler_wickets_before = lag(
      cumsum(
        wicket_type %in% c(
          "caught",
          "bowled",
          "lbw",
          "caught and bowled",
          "stumped",
          "hit wicket")),
      default = 0)) |>
  ungroup()
```

1. Batter match level stats

```{r}
batsman_match_stats <- bbl_all |>
  group_by(match_id, start_date, striker) |>
  summarise(
    
    # Total runs scored by the batter in the match
    runs_scored = sum(runs_off_bat),
    
    # Balls faced: wides do not count
    balls_faced = sum(ifelse(is.na(wides), 1, 0)),
    
    # Dismissals of this batter
    dismissals = sum(player_dismissed == striker, na.rm = TRUE),
    
    # Boundary runs scored
    boundary_runs = sum(boundary_runs),
    
    .groups = "drop"
  )

```

2. Bowler match stats
```{r}
bowler_match_stats <- bbl_all |>
  group_by(match_id, start_date, bowler) |>
  summarise(
    balls_bowled   = sum(legal_ball),
    runs_conceded  = sum(runs_off_bat + coalesce(wides, 0) + coalesce(noballs, 0)),
    wickets_taken  = sum(wicket_type %in% c("caught", "bowled", "lbw",
                                             "caught and bowled", "stumped", "hit wicket")),
    dot_balls      = sum(dot_ball),
    .groups = "drop"
  )
```


Batsman historical/career data
```{r}
batsman_history <- batsman_match_stats |>
  arrange(striker, start_date) |>
  group_by(striker) |>
  mutate(
    career_matches_before      = row_number() - 1,
    career_runs_before         = lag(cumsum(runs_scored), default = 0),
    career_balls_before        = lag(cumsum(balls_faced), default = 0),
    career_dismissals_before   = lag(cumsum(dismissals), default = 0),
    career_boundary_runs_before = lag(cumsum(boundary_runs), default = 0),

    career_batting_avg_before = ifelse(career_dismissals_before > 0,
                                        career_runs_before / career_dismissals_before,
                                        NA_real_),
    career_strike_rate_before = ifelse(career_balls_before > 0,
                                        career_runs_before / career_balls_before * 100,
                                        NA_real_),
    career_boundary_p_before  = ifelse(career_runs_before > 0,
                                        career_boundary_runs_before / career_runs_before * 100,
                                        NA_real_)
  ) |>
  ungroup() |>
  select(match_id, striker, starts_with("career_"))
```

Bowler historical/career level data


```{r}
bowler_history <- bowler_match_stats |>
  arrange(bowler, start_date) |>
  group_by(bowler) |>
  mutate(
    career_matches_bowled_before = row_number() - 1,
    career_balls_bowled_before   = lag(cumsum(balls_bowled), default = 0),
    career_runs_conceded_before  = lag(cumsum(runs_conceded), default = 0),
    career_wickets_before        = lag(cumsum(wickets_taken), default = 0),
    career_dots_before           = lag(cumsum(dot_balls), default = 0),

    career_economy_before = ifelse(career_balls_bowled_before > 0,
                                    career_runs_conceded_before / career_balls_bowled_before * 6,
                                    NA_real_),
    career_bowling_avg_before = ifelse(career_wickets_before > 0,
                                        career_runs_conceded_before / career_wickets_before,
                                        NA_real_),
    career_strike_rate_before = ifelse(career_wickets_before > 0,
                                        career_balls_bowled_before / career_wickets_before,
                                        NA_real_),
    career_dot_p_before = ifelse(career_balls_bowled_before > 0,
                                  career_dots_before / career_balls_bowled_before * 100,
                                  NA_real_)
  ) |>
  ungroup() |>
  select(match_id, bowler, starts_with("career_"))
```