---
title: "Pitch Presentation"
format: html
---

```{r}
#| echo = FALSE

knitr::opts_chunk$set(warning= FALSE, message = FALSE, echo = FALSE)
```
```{r}
library(tidyverse)
library(here)
library(yardstick)
library(xgboost)

```


```{r}
results <- readRDS(
  here("Pranav", "data", "model_2_results.rds")
)

t20_final <- results$model
t20_cv <- results$cv
test_data <- results$test_data
pred_probs <- results$pred_probs
y_test <- results$y_test
pr_auc <- results$pr_auc
roc_auc <- results$roc_auc
```

## Can we build a model that predicts the probability of a wicket falling on any given delivery, using match-state and player-identity features?


### Feature engineering
The variables engineereed for the model are:

#### Match state:

  "innings", "over_no", "phase", "balls_remaining",
  "score_before", "wickets_lost_before", "required_runs", 
  "free_hit",

#### Bowler in-match form:

  "bowler_balls_bowled_before", "bowler_economy_before",
  "bowler_boundary_p_before", "bowler_dot_ball_p_before",

#### Batsman in-match form:

  "batsman_runs_scored_before", "batsman_balls_faced_before",
  "batsman_strike_rate_before", "batsman_boundary_p_before",

#### Bistorical / career form:

  "career_matches_before", "career_batting_avg_before",
  "career_bowling_strike_rate_before","career_batting_strike_rate_before", "career_boundary_p_before",
  "career_matches_bowled_before", "career_economy_before",
  "career_bowling_avg_before", "career_dot_p_before"


### Training and test split
```r
# Identify the chronological order of matches
match_dates <- model_data |>
  distinct(match_id, start_date) |>
  arrange(start_date, match_id)

# Use the first 80% of matches chronologically as the training period
cutoff_index <- floor(nrow(match_dates) * 0.8)

# Set the date of the 80th percentile match as the cutoff
cutoff_date <- match_dates$start_date[cutoff_index]

# Train the model only on matches occurring before or on the cutoff date
train_data <- model_data |>
  filter(start_date <= cutoff_date)

# Test the model on the most recent 20% of matches
test_data <- model_data |>
  filter(start_date > cutoff_date)
```
### Model
An XGBoost model was used on the data. This initial model had a PR-AUC score of 0.096, and an ROC-AUC score of 0.623. 

### Feature importance

```{r}
importance <- xgb.importance(
  model = t20_final
)

importance |>
  slice_max(
    order_by = Gain,
    n = 10
  ) |>
  ggplot(
    aes(
      x = reorder(Feature, Gain),
      y = Gain
    )
  ) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Predictors",
    x = NULL,
    y = "XGBoost Gain"
  ) +
  theme_minimal()
```

This shows that the initial iterations of the model are giving more importance to match-level variables and not career-level variables. 

### Next Steps
1. Tune and validate the wicket model : Test different XGBoost parameters and compare performance against simpler baseline models.
2. Analyse model errors : Identify match situations where the model struggles to predict wickets.
3. Answer the next question : "Which bowlers are most effective at restricting runs and taking wickets in the death overs, after controlling for the match state they bowl in?"